---
title: 'User Behaviour Analysis & Churn Propensity (SaaS Domain)'
author: "Malika Diwakar"
date: "`r lubridate::now(tzone = 'Asia/Kolkata')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: yeti
    highlight: zenburn
    df_print: paged
    code_folding: show
  word_document:
    toc: true
    toc_depth: '2'
  pdf_document:
    toc: true
    toc_depth: '2'
editor_options:
  chunk_output_type: console
---

#### Global Settings 

```{r setup, include = FALSE}

# specify global settings for all code chunks 
knitr::opts_chunk$set(
  # print code for subsequent code chunks 
  echo = TRUE,
  # do not print messages 
  message = FALSE,
  # do not print warnings 
  warning = FALSE)

```

#### Load Libraries

```{r libraries}

library(dplyr)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(ggthemes)
library(plotly)
library(scales)
library(data.table)
library(tidyr)
library(janitor)
library(skimr)
library(caret)
library(randomForest)
library(xgboost)
library(pROC)
library(ROCR)
library(e1071)

```

---

## Data Preparation & Feature Engineering

```{r prep}

# set working directory 
setwd("/Users/malika.diwakar/Desktop/Churn Analysis Project")

## import dataset 
dataset <- read.csv("ChurnAnalysisData.csv", header = TRUE)

# view data structure 
str(dataset)

# change date columns to DATE type 
## active_month originally contains only "YYYY-MM" (therefore '-01' is added only to enable date conversion in R) 
dataset$active_month <- as.Date(paste0(dataset$active_month, "-01"))
## first_payment_date 
dataset$first_payment_date <- as.Date(dataset$first_payment_date)

# confirm changes have been made 
str(dataset)

# remove serial no. column (not useful for analysis) 
dataset <- dataset %>% 
  select(-X)

# summary statistics 
summary(dataset)

# check for missing values 
dataset %>%
  summarise(across(everything(), ~sum(is.na(.))))

# sort each user's data in the order of time & label the tenure months 
dataset <- dataset %>% 
  group_by(user_id) %>% 
  arrange(active_month, .by_group=TRUE) %>% 
  mutate(
    tenure_month = row_number()
  )

# create a binary "churn indicator" (where 1=churn, 0=no churn) 
## If user stays till next month -> return value of 0 
## If user does not stay till next month -> return value of 1 
dataset <- dataset %>% 
  group_by(user_id) %>% 
  mutate(
    next_month = lead(active_month),
    churn_indicator = ifelse(is.na(next_month), 1, 0)
  )

# create binary variables - was the specific feature used at all? (1=yes, 0=no) 
dataset <- dataset %>% 
  mutate(
    used_cull = ifelse(culls > 0, 1, 0),
    used_edit = ifelse(edits > 0, 1, 0),
    used_retouch = ifelse(retouchs > 0, 1, 0)
  )

# create variables "prev_projects" & "project_drop" to check for decline in usage 
dataset <- dataset %>%
  group_by(user_id) %>%
  arrange(active_month, .by_group = TRUE) %>%
  mutate(
    prev_projects = lag(total_projects),
    project_drop = total_projects - prev_projects
  ) %>%
  ungroup()

# view section of dataset after making above changes 
head(dataset)

```

**Interpretation:**

- Converted date columns into appropriate formats for time-based analysis. 
- Removed columns irrelevant to the analysis. 
- Created new variables: 
  - `tenure_month` - to track user lifecycle 
  - `churn_indicator` - to classify whether or not users churned (where 1 = churn, and 0 = no churn) 
  - `used_cull`, `used_edit`, and `used_retouch` - binary indicators to capture feature usage 
  - `project_drop` - to measure month-over-month changes in engagement 

---

## Data Validation

```{r validation}

# since active_month originally contained only YYYY-MM values, '-01' was appended only to enable DATE conversion in R 

# check for cases where activity month occurs before payment month 
true_mismatch <- dataset %>%
  filter(
    format(active_month, "%Y-%m") <
    format(first_payment_date, "%Y-%m")
  )

# count mismatched rows 
nrow(true_mismatch)

# count affected users 
n_distinct(true_mismatch$user_id)

# calculate % of mismatched rows 
round((nrow(true_mismatch) / nrow(dataset)) * 100, 2)

```

**Notes:**

- `active_month`was originally provided only at month-level granularity (YYYY-MM). To enable time-based analysis, `-01` was appended during date conversion. Therefore, this field should be interpreted as representing an *activity month rather than an exact activity date*. 
- Some observations initially appeared to show activity occurring before the recorded `first_payment_date.` However, many of these cases were due to *differences in granularity between month-level activity data and exact payment dates*. 
- After validating at the year-month (Y-m) level, `only 0.35%` of observations showed activity occurring before the payment month itself - a very small proportion of the dataset that is unlikely to materially affect the overall analysis. 
- These remaining cases may reflect: 
  - free trial usage, 
  - delayed billing recognition, 
  - or differences in business logic between activity and payment systems. 

Since the analysis primarily relies on observed user activity patterns rather than payment timing, the impact on churn and engagement insights is expected to be minimal. 

---

## Analysis: Overall User Behaviour

```{r overall}

# overall user base behaviour based on projects made & features used 
overall_behaviour <- dataset %>%
  summarise(
    # distinct no. of subscribers 
    distinct_users = n_distinct(user_id),
    # average projects per month 
    avg_projects = round(mean(total_projects), 1),
    # average culls per month 
    avg_culls = round(mean(culls), 1),
    # average edits per month 
    avg_edits = round(mean(edits), 1),
    # average retouches per month 
    avg_retouchs = round(mean(retouchs), 1)
  )

overall_behaviour

```

**Interpretation:**

- There are `12,175 distinct` users in the dataset. 
- On average, users create `8.8 projects` per active month. 

Feature usage patterns: 

- `Editing` is the most widely used feature (`~5.6` edits per month). 
- `Culling` follows closely behind (`~4.7` culls per month). 
- `Retouching` is significantly less used (`~0.3` retouches per month) due to its recent introduction on the platform. This average is weighed down by users who haven't encountered the feature yet. 

_Editing and Culling are central to user workflows, while Retouching is still yet to gain traction as more users discover the feature._ 

---

## Churned vs Retained Subscribers

```{r churn}

# check for changes in user behaviour before churn 
## churned vs retained: average no. of projects & features used 
churn_analysis <- dataset %>%
  group_by(churn_indicator) %>%
  summarise(
    avg_projects = mean(total_projects),
    avg_culls = mean(culls),
    avg_edits = mean(edits),
    avg_retouch = mean(retouchs)
  )

churn_analysis

## churned vs retained: check % of total users that used each feature 
feature_usage_comparison <- dataset %>%
  group_by(churn_indicator) %>%
  summarise(
    used_cull = mean(used_cull),
    used_edit = mean(used_edit),
    used_retouch = mean(used_retouch)
  )

feature_usage_comparison

# plot feature usage graph 
## step 1: clean feature names 
feature_plot <- feature_usage_comparison %>%
  pivot_longer(
    cols = c(used_cull, used_edit, used_retouch),
    names_to = "feature",
    values_to = "usage"
  ) %>%
  mutate(
    feature = recode(feature,
      used_cull = "Used Culling",
      used_edit = "Used Editing",
      used_retouch = "Used Retouching"
    )
  )

# relabel "0" and "1" in legend 
feature_plot <- feature_plot %>%
  mutate(
    churn_indicator = factor(
      churn_indicator,
      levels = c(0, 1),
      labels = c("Retained", "Churned")
    )
  )

## step 2: plot the graph 
ggplotly(ggplot(feature_plot, aes(x = feature, y = usage, fill = factor(churn_indicator))) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c("#2E86AB", "#F18F01"),
                    labels = c("Retained", "Churned")) +
  labs(
    title = "Feature Usage: Churned vs Retained Users",
    x = "Feature",
    y = "Usage Rate",
    fill = "User Type"
  ) +
  theme_minimal()
)

```

**Interpretation:**

- Churned users show a modest decline in activity, indicating that _churn is not driven by a sharp drop in usage_. 
- `Editing` shows the _largest usage gap_ (retained users use editing ~9% more), suggesting it to be a key retention driver. 
- Higher `retouch` usage among churned users indicates _isolated feature usage_ rather than full product adoption (culling -> editing -> retouching). 

---

## Month-on-Month (MoM) - Time Series Analysis

```{r mom_1}

# churn rate by each tenure month 
churn_by_tenure <- dataset %>%
  group_by(user_id) %>%
  summarise(tenure_month = max(tenure_month)) %>% # when each user churned 
  group_by(tenure_month) %>%
  summarise(churn_count = n()) %>%
  mutate(
    churn_rate = churn_count / sum(churn_count)
  )

# plot the churn rate graph 
## step 1: clean values 
churn_by_tenure <- churn_by_tenure %>%
  mutate(churn_rate = round(churn_rate, 3))
## step 2: plot the graph 
ggplotly(ggplot(churn_by_tenure, aes(x = tenure_month, y = churn_rate)) +
    geom_line(color="#B22222") +
    geom_point(color="#1874CD") +
    scale_y_continuous(
      labels = scales::label_percent(accuracy = 1),
      limits = c(0, 0.2),
      breaks = seq(0, 0.2, by = 0.05)
    ) +
    scale_x_continuous(
      breaks = seq(0, 25, by = 5),
      limits = c(0, 25)
    ) +
    labs(
      title = "Churn Distribution by Tenure Month (%)",
      x = "Tenure Month",
      y = "% of Total Churn"
    ) +
    theme_minimal()
)

```

**Interpretation:** 

The highest churn activity occurs within the first 5 months, highlighting early-stage risk. Early engagement and feature adoption is critical for long-term subscriber retention. 

```{r mom_2}

# total projects MoM 
mom_projects <- dataset %>%
  group_by(tenure_month) %>%
  summarise(
    avg_projects = mean(total_projects)
  )

# plot the average MoM projects graph 
ggplotly(ggplot(mom_projects, aes(x=tenure_month, y=avg_projects)) +
  geom_line(color="#009ACD") +
  geom_point(color="#CD5B45") +
  labs(
    title = "Projects Completed Over the Tenure Months",
    x = "Tenure Month",
    y = "Average Number of Projects"
  ) +
  theme_minimal()
)

```

**Interpretation:** 

Users who continue their subscriptions for longer periods display increased project activity while early-stage users show lower/declining usage. 

```{r mom_3}

# MoM feature usage 
mom_feature_usage <- dataset %>%
  group_by(tenure_month) %>%
  summarise(
    avg_projects = mean(total_projects),
    avg_culls = mean(culls),
    avg_edits = mean(edits),
    avg_retouch = mean(retouchs)
  )

# plot the MoM feature usage graph - multi-line 
## step 1: clean feature names 
mom_feature_plot <- mom_feature_usage %>%
  pivot_longer(
    cols = c(avg_projects, avg_culls, avg_edits, avg_retouch),
    names_to = "feature",
    values_to = "value"
  ) %>%
  mutate(
    feature = recode(feature,
      avg_projects = "Projects",
      avg_culls = "Culling",
      avg_edits = "Editing",
      avg_retouch = "Retouching"
    )
  )

## step 2: create the plot 
ggplotly(ggplot(mom_feature_plot, aes(x = tenure_month, y = value, color = feature)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.7) +
  scale_color_manual(values = c("#1B9E77", "#D95F02", "#7570B3", "#E7298A")) +
  labs(
    title = "How Feature Usage Has Evolved Over Time",
    x = "Tenure Month",
    y = "Average Usage",
    color = "Feature"
  ) +
  theme_minimal()
)

```

**Interpretation:** 

`Editing` and `Culling` contribute the most to overall projects (and are often used together), while `Retouching` remains the least used feature mainly because it was recently introduced. Among established users, however, average `Retouching` usage has already grown from ~0.5 to ~1.3 per month over 22 active months, suggesting strong adoption momentum. 

---

## High-Risk Target Group/Segment (TG)

```{r tg_1}

# define threshold (bottom 25% quartile) for total projects done by TG 
threshold_projects <- quantile(dataset$total_projects, 0.25, na.rm = TRUE)

# classification: TG vs non-TG 
dataset <- dataset %>%
  mutate(
    high_risk_flag = ifelse(
      used_edit == 0 &
      tenure_month <= 5 &
      total_projects <= threshold_projects &
      (!is.na(project_drop) & project_drop < 0),
      1, 0
    )
  )

```

**Interpretation:** 

Based on the analysis, the high-risk target segment can be defined by: 

- No editing usage (i.e. not using the core popular feature) 
- Early lifecycle stage (tenure month <= 5) 
- Low project activity (bottom 25% quartile) 
- Declining engagement (project_drop < 0) 

_Users in this segment are not receiving enough value from the product, making them highly prone to churning._ 

```{r tg_2}

# extract only user-level data (total 12,175 distinct users) 
user_level_data <- dataset %>%
  group_by(user_id) %>%
  summarise(
    high_risk_flag = max(high_risk_flag),
    churn_rate = mean(churn_indicator),
    avg_projects = mean(total_projects),
    avg_culls = mean(culls),
    avg_edits = mean(edits),
    avg_retouch = mean(retouchs),
    edit_adoption = mean(used_edit),
    cull_adoption = mean(used_cull),
    retouch_adoption = mean(used_retouch)
  )

# what % of total users are from the target segment? 
tg_size <- user_level_data %>%
  ungroup() %>% 
  summarise(
    total_users = n_distinct(user_id),
    tg_users = n_distinct(user_id[high_risk_flag == 1])
  ) %>%
  mutate(
    tg_percentage = tg_users / total_users
  )

tg_size

```

**Interpretation:** 

Out of 12,175 total users, *~30% users (3,613)* are part of the high-risk segment. 

``` {r tg_3}

# quantify & compare TG vs non-TG 
tg_compare <- user_level_data %>%
  group_by(high_risk_flag) %>%
  summarise(
    users = n(),
    churn_rate = mean(churn_rate),
    avg_projects = mean(avg_projects),
    avg_culls = mean(avg_culls),
    avg_edits = mean(avg_edits),
    avg_retouch = mean(avg_retouch)
  ) %>%
  mutate(
    segment = ifelse(high_risk_flag == 1, "High Risk TG", "Others"),
    user_percentage = users / sum(users)
  )

tg_compare

# plot TG vs non-TG churn rate graph 
## step 1: clean feature names 
tg_plot <- user_level_data %>%
  group_by(high_risk_flag) %>%
  summarise(
    churn_rate = mean(churn_rate)
  ) %>%
  mutate(
    segment = ifelse(high_risk_flag == 1, "High Risk", "Others")
  )

## step 2: create the plot 
ggplotly(ggplot(tg_plot, aes(x = segment, y = churn_rate, fill = segment)) +
  geom_col(width = 0.6) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c("#E63946", "#457B9D")) +
  labs(
    title = "Churn Rate: High-Risk vs Other Users",
    x = "",
    y = "Churn Rate",
    fill = "Segment"
  ) +
  theme_minimal()
)

```

**Interpretation:** 

The target group (TG) is a sizable `30% of the user base`. 

Now that we have a defined TG and a comparison of usage patterns, we can see that: 

- Feature usage and project activity is significantly lower in the TG, indicating low engagement across the full editing workflow. 
- However, the non-TG shows a higher churn rate (45%) than the TG's (23%), suggesting that churn is not solely driven by low usage, but may also be influenced by alternative behaviours such as: 
  - completion of user objectives 
  - switching to alternatives that suit their needs better 
  - pricing factors 

---

## Retention Curve 

```{r rc}

# show distinct active users across each month 
retention_curve <- dataset %>%
  group_by(tenure_month) %>%
  summarise(
    active_users = n_distinct(user_id)
  )

# plot the retention curve 
ggplotly(ggplot(retention_curve, aes(x = tenure_month, y = active_users)) +
  geom_line(linewidth = 0.8, color = "#E63946") +
  geom_point(size = 1.7, color = "#1D3557") +
  labs(
    title = "User Retention Curve",
    x = "Tenure Month",
    y = "Active Users"
  ) +
  theme_minimal()
)

```

**Interpretation:** 

Retention drops significantly in months 2 (`-2,290 users`) and 3 (`-2,099 users`), highlighting the importance of early engagement in reducing churn. Survival past this window is a strong signal of product-value fit, correlating with markedly higher long-term retention. 

---

## Predictive Analysis

### Redefining Churn Using a Cutoff Month

```{r pred_1}

# define cutoff as the most recent month in the dataset 
cutoff_month <- max(dataset$active_month)

# create a true user-level churn flag 
user_churn <- dataset %>%
  group_by(user_id) %>%
  summarise(last_active_month = max(active_month)) %>%
  mutate(
    # if last active month is before the cutoff, user has churned 
    churned = ifelse(last_active_month < cutoff_month, 1, 0)
  ) %>%
  select(user_id, churned)

# check class balance 
table(user_churn$churned)

prop.table(table(user_churn$churned))

# each user should appear as ONE row with aggregated behavioural features 
# target variable is whether the user ultimately churned or not 
model_data <- dataset %>%
  group_by(user_id) %>%
  summarise(
    total_tenure     = max(tenure_month),
    avg_projects     = mean(total_projects),
    avg_culls        = mean(culls),
    avg_edits        = mean(edits),
    avg_retouch      = mean(retouchs),
    edit_adoption    = mean(used_edit),
    cull_adoption    = mean(used_cull),
    retouch_adoption = mean(used_retouch),
    avg_project_drop = mean(project_drop, na.rm = TRUE),
    high_risk_flag   = max(high_risk_flag)
  ) %>%
  ungroup() %>%
  # join the newly defined churn flag 
  left_join(user_churn, by = "user_id")

# confirm class balance is consistent 
table(model_data$churned)

prop.table(table(model_data$churned))

```

### Prepare Features for Modelling

```{r pred_2}

# remove user_id (not a predictor) and convert target variable to factor for classification 
model_df <- model_data %>%
  select(-user_id) %>%
  mutate(churned = factor(churned, levels = c(0, 1), labels = c("Retained", "Churned")))

# check for any remaining missing values 
colSums(is.na(model_df))

# replace any missing values in avg_project_drop with 0 (i.e. when first month has no lag) 
model_df <- model_df %>%
  mutate(avg_project_drop = replace_na(avg_project_drop, 0))

# confirm there are no more missing values 
colSums(is.na(model_df))

```

### Split Training & Testing Data

```{r pred_3}

# set seed for reproducibility 
set.seed(123)

# split training & testing data in 80-20 partition 
train_index <- createDataPartition(model_df$churned, p = 0.8, list = FALSE)

train_data <- model_df[train_index, ]

test_data  <- model_df[-train_index, ]

# confirm class balance is maintained in both splits 
prop.table(table(train_data$churned))

prop.table(table(test_data$churned))

```

### Define Cross-Validation Control

Using 5-fold cross-validation splits the training data into 5 parts (trains on 4, validates on 1, and repeats 5 times), giving a more reliable performance estimate than a single train/test split. 

```{r pred_4}

# use 5-fold cross-validation during training to reduce overfitting 
train_control <- trainControl(
  method = "cv",                      # cross-validation 
  number = 5,                         # 5 folds 
  classProbs = TRUE,                  # return predicted probabilities (needed for AUC/ROC metrics) 
  summaryFunction = twoClassSummary,  # returns AUC, Sensitivity, Specificity 
  savePredictions = "final"
)

```

## Model 1 - Logistic Regression

Logistic Regression: a linear model that estimates the probability of churn, (a good baseline for binary classification). 

```{r pred_5_logit}

# set seed for reproducibility 
set.seed(123)

# build logit model 
model_logit <- train(
  churned ~ .,
  data = train_data,
  method = "glm",
  family = "binomial",
  metric = "ROC",
  trControl = train_control
)

# view logistic model summary 
summary(model_logit)

# predict on test data 
pred_logit <- predict(model_logit, newdata = test_data)

pred_logit_prob <- predict(model_logit, newdata = test_data, type = "prob")[, "Churned"]

# confusion matrix 
cm_logit <- confusionMatrix(pred_logit, test_data$churned, positive = "Churned")

cm_logit

# AUC-ROC: measures how well the model distinguishes churned vs retained 
roc_logit <- roc(test_data$churned, pred_logit_prob, levels = c("Retained", "Churned"))

auc(roc_logit)

```

**Interpretation:** 

**1. Model Coefficients & Predictor Significance** 

- All 11 predictors are statistically significant, with 10 significant at the 99.9% confidence level (3 stars *** ) and `cull_adoption` significant at the 95% level (1 star * ). 
- Negative coefficients indicate the variable reduces churn probability. 
  - `retouch_adoption` has the strongest protective effect (-2.15), followed by `edit_adoption` (-0.63), 
  - suggesting that consistent feature adoption is the biggest driver of retention. 
- Positive coefficients indicate the variable increases churn probability. 
  - `avg_projects` (+0.13) and `high_risk_flag` (+0.24) both increase churn risk, 
  - confirming that the high-risk segment defined in the EDA is a valid churn signal. 
- `total_tenure` (-0.14) confirms that the longer a user stays, the less likely they are to churn. 
  - This is consistent with the EDA finding that survival past the first 5 months strongly predicts long-term retention. 

**2. Model Fit** 

The model explains a meaningful portion of churn variance (null deviance: 13,354 and residual deviance: 11,374), performing reasonably well as a baseline with an AUC of 0.74. 

**3. Limitation** 

Assumption of linear relationships: non-linear patterns in user behaviour are not captured here, which is where Random Forest and XGBoost are expected to improve. 

---

## Model 2 - Random Forest

Random Forest: an ensemble of decision trees. 

- Each tree is trained on a random subset of data and features. 
- Final prediction = majority vote across all trees. 
- More powerful than logistic regression for non-linear relationships. 

```{r pred_6_rf}

# set seed for reproducibility 
set.seed(123)

# build random forest model 
model_rf <- train(
  churned ~ .,
  data = train_data,
  method = "rf",
  metric = "ROC",
  trControl = train_control,
  tuneLength = 3
)

# view tuning results 
print(model_rf)

# feature importance in driving churn 
varImp(model_rf) %>% plot(main = "Random Forest - Feature Importance")

# predict on test data 
pred_rf <- predict(model_rf, newdata = test_data)

pred_rf_prob <- predict(model_rf, newdata = test_data, type = "prob")[, "Churned"]

# confusion matrix 
cm_rf <- confusionMatrix(pred_rf, test_data$churned, positive = "Churned")

cm_rf

# AUC-ROC 
roc_rf <- roc(test_data$churned, pred_rf_prob, levels = c("Retained", "Churned"))

auc(roc_rf)

```

**Interpretation:** 

**1. Hyperparameter Tuning** 

- The model was tested across 3 values of `mtry` (the number of features randomly considered at each tree split): 2, 6, and 10. 
- `mtry = 2` was selected as optimal based on the highest AUC/ROC (0.766), despite higher `mtry` values producing slightly better sensitivity. 

**2. Feature Importance** 

- `total_tenure` (100) is the strongest predictor of churn, followed closely by `avg_edits` (97.98), `avg_projects` (96.57), and `avg_project_drop` (95.13), confirming that lifecycle stage and engagement trends are the most predictive signals. 
- `high_risk_flag` scored 0, meaning it adds no additional predictive value beyond its underlying features that are are already included in the model (as expected). 

**3. Model Performance** 

- AUC = 0.7579, a slight improvement from the logistic regression model (0.7403), indicating that Random Forest better captures non-linear relationships in the data. 
- Accuracy improved slightly to 69.8% (vs 68.5% in logistic regression). 
- Sensitivity held steady at 68.3%, identifying the same proportion of actual churners as logistic regression. 
- Specificity improved to 71.1% (vs 68.7% in logistic regression), suggesting that Random Forest is better at correctly identifying retained users, which helps reduce unnecessary retention spend on low-risk users. 

---

## Model 3 - XGBoost

XGBoost: gradient-boosted trees, where each new tree corrects the errors of the previous one. 

```{r pred_7_xgb}

# XGBoost requires a numeric matrix for predictors and numeric 0/1 values for the target variable 
# prepare training and testing data (keep churned as factor for later evaluation) 
train_xgb <- train_data %>%
  mutate(churned = factor(churned, levels = c("Retained", "Churned")))

test_xgb <- test_data %>%
  mutate(churned = factor(churned, levels = c("Retained", "Churned")))

# prepare training matrix 
train_matrix <- xgb.DMatrix(
  data = as.matrix(train_xgb %>% select(-churned)),
  label = ifelse(train_xgb$churned == "Churned", 1, 0)
)

# prepare test matrix 
test_matrix <- xgb.DMatrix(
  data = as.matrix(test_xgb %>% select(-churned)),
  label = ifelse(test_xgb$churned == "Churned", 1, 0)
)

# set parameters 
xgb_params <- list(
  objective = "binary:logistic",
  eval_metric = "auc",
  eta = 0.1,
  max_depth = 5,
  subsample = 0.8,
  colsample_bytree = 0.8,
  min_child_weight = 1,
  gamma = 0
)

# set seed for reproducibility 
set.seed(123)

# train with cross-validation to find best number of rounds 
xgb_cv <- xgb.cv(
  params = xgb_params,
  data = train_matrix,
  nrounds = 200,
  nfold = 5,
  verbose = 0,
  early_stopping_rounds = 20
)

# use best_iteration if available, otherwise use round with highest test AUC 
best_nrounds <- if (!is.null(xgb_cv$best_iteration) && length(xgb_cv$best_iteration) > 0) {
  xgb_cv$best_iteration
} else {
  which.max(xgb_cv$evaluation_log$test_auc_mean)
}

best_nrounds

# train final model using best number of rounds 
model_xgb <- xgb.train(
  params = xgb_params,
  data = train_matrix,
  nrounds = best_nrounds,
  verbose = 0
)

# feature importance 
xgb_importance <- xgb.importance(model = model_xgb)

xgb.plot.importance(xgb_importance, main = "XGBoost - Feature Importance")

# predict on test set 
pred_xgb_prob <- predict(model_xgb, newdata = test_matrix)

# convert probabilities to class labels using 0.5 threshold 
pred_xgb <- factor(
  ifelse(pred_xgb_prob >= 0.5, "Churned", "Retained"),
  levels = c("Retained", "Churned")
)

# confusion matrix 
cm_xgb <- confusionMatrix(pred_xgb, test_xgb$churned, positive = "Churned")

cm_xgb

# AUC-ROC 
roc_xgb <- roc(test_xgb$churned, pred_xgb_prob, levels = c("Retained", "Churned"))

auc(roc_xgb)

```

**Interpretation:** 

**1. Training & Tuning** 

86 boosting rounds were identified as optimal, meaning the model built 86 sequential trees, each correcting the errors of the previous one, before performance stopped improving. 

**2. Feature Importance** 

- `avg_retouch` (0.236) and `total_tenure` (0.216) are the top 2 predictors, a slight change from Random Forest where `total_tenure` ranked first, suggesting XGBoost is picking up on retouching behaviour as a stronger differentiating signal than the other models. 
- `avg_edits`, `avg_projects`, and `avg_project_drop` continue to rank highly, consistent with findings across all 3 models. 
- `high_risk_flag` again scores near zero (0.005), further confirming it carries no independent predictive value beyond its underlying features. 

**3. Model Performance** 

- AUC = 0.7796 (highest across all 3 models) confirms that XGBoost best captures the complexity of churn behaviour in this dataset. 
- Accuracy improved to 70.9%, sensitivity to 70%, and specificity to 71.6%, all being improvements over both previous models. 
- The model is well-balanced between catching actual churners (70% sensitivity) and correctly clearing retained users (71.6% specificity), making it the most reliable model for risk scoring. 

---

## Model Comparison

```{r compare_models}

# compile AUC scores for all 3 models 
model_comparison <- data.frame(
  Model = c("Logistic Regression", "Random Forest", "XGBoost"),
  AUC = round(c(
    auc(roc_logit),
    auc(roc_rf),
    auc(roc_xgb)
  ), 4),
  Accuracy = round(c(
    cm_logit$overall["Accuracy"],
    cm_rf$overall["Accuracy"],
    cm_xgb$overall["Accuracy"]
  ), 4),
  Sensitivity = round(c(
    cm_logit$byClass["Sensitivity"],
    cm_rf$byClass["Sensitivity"],
    cm_xgb$byClass["Sensitivity"]
  ), 4),
  Specificity = round(c(
    cm_logit$byClass["Specificity"],
    cm_rf$byClass["Specificity"],
    cm_xgb$byClass["Specificity"]
  ), 4)
)

model_comparison

# plot ROC curves for all 3 models 
## step 1: combine all 3 ROC objects into a named list 
roc_list <- list(
  "Logistic Regression" = roc_logit,
  "Random Forest" = roc_rf,
  "XGBoost" = roc_xgb
)
## step 2: plot all ROC curves together 
ggplotly(
  ggroc(roc_list, linewidth = 0.8) +
    scale_color_manual(
      values = c(
        "Logistic Regression" = "#457B9D",
        "Random Forest" = "#2A9D8F",
        "XGBoost" = "#E63946"
      ),
      labels = c(
        paste("Logistic Regression (AUC =", round(auc(roc_logit), 3), ")"),
        paste("Random Forest (AUC =", round(auc(roc_rf), 3), ")"),
        paste("XGBoost (AUC =", round(auc(roc_xgb), 3), ")")
      )
    ) +
    geom_abline(slope = 1, intercept = 1,
                linetype = "dashed", color = "grey50") +
    labs(
      title = "ROC Curve Comparison - All Models",
      x = "Specificity",
      y = "Sensitivity",
      color = "Model"
    ) +
    theme_minimal()
)

```

**Interpretation:** 

- All 3 models perform meaningfully above random chance, with AUC scores ranging between 0.74 - 0.78. 
- XGBoost is the best performing model across every metric - highest AUC (0.7796), accuracy (70.9%), sensitivity (70%), and specificity (71.6%) - and is therefore selected as the final model for risk tier stratification. 

---

## Risk Tier Stratification using Best Model

We will use XGBoost (the best performing model). 

```{r risk_tiers}

# prepare full user-level dataset as a numeric matrix 
full_model_df <- model_data %>%
  select(-user_id, -churned) %>%
  mutate(avg_project_drop = replace_na(avg_project_drop, 0))

full_matrix <- xgb.DMatrix(
  data = as.matrix(full_model_df)
)

# generate churn probability scores for every user using XGBoost 
churn_probs <- predict(model_xgb, newdata = full_matrix)

# assign risk tiers using 33rd / 67th percentile cut-offs (dividing into thirds) 
risk_scores <- model_data %>%
  select(user_id, churned) %>%
  mutate(
    churn_prob = churn_probs,
    risk_tier = case_when(
      churn_prob >= quantile(churn_probs, 0.67) ~ "High Risk",
      churn_prob >= quantile(churn_probs, 0.33) ~ "Medium Risk",
      TRUE ~ "Low Risk"
    ),
    risk_tier = factor(risk_tier, levels = c("Low Risk", "Medium Risk", "High Risk"))
  )

# distribution of users across tiers 
risk_scores %>%
  count(risk_tier) %>%
  mutate(percentage = scales::percent(n / sum(n)))

```

**Interpretation:** 

- XGBoost was used to generate individual churn probability scores between 0 and 1 for all 12,175 users. 
- Users were then divided into 3 risk tiers using the 33rd and 67th percentiles. 
- The 3 tiers are broadly even in size: 
  - Low Risk (33%, 4,012 users) 
  - Medium Risk (34%, 4,145 users) 
  - High Risk (33%, 4,018 users) 

- The even distribution is intentional by splitting at percentile thresholds rather than fixed probability values. 
- Every user is assigned a tier relative to the rest of the user base, ensuring no tier is too small to act on or too large to target efficiently. 

--- 

## Risk Tier Profiling & Benchmarking

```{r}

# join risk scores with full behavioural features for profiling 
risk_profile <- risk_scores %>%
  left_join(model_data, by = "user_id")

# behavioural summary per tier 
tier_summary <- risk_profile %>%
  group_by(risk_tier) %>%
  summarise(
    users = n(),
    actual_churn_rate = scales::percent(mean(churned.x), accuracy = 0.1),
    avg_churn_prob = round(mean(churn_prob), 3),
    avg_tenure = round(mean(total_tenure), 1),
    avg_projects = round(mean(avg_projects), 1),
    avg_edits = round(mean(avg_edits), 1),
    edit_adoption_pct = scales::percent(mean(edit_adoption), accuracy = 0.1),
    cull_adoption_pct = scales::percent(mean(cull_adoption), accuracy = 0.1),
    high_risk_pct = scales::percent(mean(high_risk_flag), accuracy = 0.1)
  )

tier_summary

# plot churn probability distribution by risk tier 
ggplotly(
  ggplot(risk_scores, aes(x = churn_prob, fill = risk_tier)) +
    geom_histogram(bins = 40, alpha = 0.8, position = "identity") +
    scale_fill_manual(values = c("#2A9D8F", "#F4A261", "#E63946")) +
    labs(
      title = "Churn Probability Distribution by Risk Tier",
      x = "Predicted Churn Probability",
      y = "Number of Users",
      fill = "Risk Tier"
    ) +
    theme_minimal()
)

# plot actual vs predicted churn rate per tier 
## step 1: separate plot data for risk tiers 
tier_plot_data <- risk_profile %>%
  group_by(risk_tier) %>%
  summarise(
    actual_churn_rate = mean(churned.x),
    avg_churn_prob = mean(churn_prob)
  ) %>%
  pivot_longer(
    cols = c(actual_churn_rate, avg_churn_prob),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = recode(metric,
      actual_churn_rate = "Actual Churn Rate",
      avg_churn_prob = "Avg Predicted Probability"
    )
  )
## step 2: create plot 
ggplotly(
  ggplot(tier_plot_data, aes(x = risk_tier, y = value, fill = metric)) +
    geom_col(position = "dodge", width = 0.8) +
    scale_y_continuous(labels = scales::percent_format()) +
    scale_fill_manual(values = c("#457B9D", "#E63946")) +
    labs(
      title = "Actual vs Predicted Churn Rate by Risk Tiers",
      x = "Risk Tier",
      y = "Rate / Probability",
      fill = "Metric"
    ) +
    theme_minimal()
)

```

**Interpretation:** 

- `Low Risk` users have: 
  - lowest churn rate (10.8%) 
  - longest average tenure (9.4 months) 
  - highest project activity (12.3 projects, 8.2 edits) 
  - strong feature adoption (86.6% editing, 73.4% culling) 
  
_These are well-engaged users who have clearly found value in the product._ 

- `Medium Risk` users sit in the middle with: 
  - 45.2% churn rate 
  - shorter tenure (4.8 months) 
  - moderately lower engagement 
  
_They are active but have not yet demonstrated the consistent feature adoption seen in Low Risk users._ 

- `High Risk` users show: 
  - 75.3% churn rate 
  - shortest tenure (3.4 months) 
  - lowest project activity (4.3 projects, 1.8 edits) 
  - significantly lower feature adoption (44.8% editing, 53.1% culling) 
  
_This is consistent with early-stage users who have not yet realised product value._ 

### Export Data to CSV

```{r export_data, include = FALSE}

# user-level risk tier assignments 
write.csv(
  risk_scores %>% select(user_id, churn_prob, risk_tier),
  "/Users/malika.diwakar/Desktop/Churn Analysis Project/user_risk_tiers.csv",
  row.names = FALSE
)

# cleaned version of original dataset 
write.csv(
  dataset, 
  "/Users/malika.diwakar/Desktop/Churn Analysis Project/cleaned_data.csv", 
  row.names = FALSE)

```

